Spatial Reasoning and Data Displays

### 2015-03-16 Paper/InfoVis2015/TestingVisualAptitude.tex
InfoVis style file submission

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

<<VisualAptitude-setup, fig.keep='all', cache=FALSE, echo=FALSE, eval=TRUE, message=F, warning=F>>=
rm(list=ls())
wd <- getwd()
library(extrafont)
library(knitr)
opts_chunk$set(fig.path='Figure/fig-', cache.path='cache/', fig.align='center', fig.width=5, fig.height=5, fig.show='hold', par=TRUE, cache=F, tidy=T, concordance=TRUE, autodep=TRUE, root.dir="../", comment="|", dev="CairoPDF")
datadir <- "../../Data/"
imgdir <- "Figure/"
codedir <- "../../Code/"

options(replace.assign=TRUE,scipen=3)

library(reshape2)
suppressMessages(library(ggplot2))
library(plyr)
library(grid)
suppressMessages(library(gridExtra))
library(xtable)
library(stringr)
library(magrittr)
source(paste(codedir, "Analysis.R", sep=""))

sanitize <- function(x){
  str_replace(gsub("_", ".", x, fixed=T), "Pr\\((.{1,}))", "Pr$(\\1)$")
}

@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

\section{Results}\label{sec:results}
Results are based on an evaluation of \Sexpr{nrow(ans)} undergraduate students at Iowa State University. \Sexpr{round(sum(ans.summary$stem)/nrow(ans.summary)*100)}\% of the participants were in STEM fields, the others were distributed relatively evenly between agriculture, business, and the social sciences. Students were evenly distributed by gender, and were between 18 and 24 years of age with only one exception. This is reasonably representative\footnote{\url{http://www.registrar.iastate.edu/sites/default/files/uploads/stats/university/F14summary.pdf}} of the university as a whole; in the fall 2014 semester, 26\% of students were associated with the college of engineering, 24\% were associated with the college of liberal arts and sciences, 15\% were associated with the college of human sciences, 7\% with the college of design, 13\% with the business school, and 15\% with the school of agriculture.

@@

\begin{table*}[htb]\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure Classification & Visual Search  \\\hline
ISU Students & \Sexpr{round(mean(ans.summary$card_rot), 1)} (\Sexpr{round(sd(ans.summary$card_rot), 1)}) 
             & \Sexpr{round(mean(ans.summary$folding), 1)} (\Sexpr{round(sd(ans.summary$folding), 1)})
             & \Sexpr{sprintf("%.1f", mean(ans.summary$fig_class))} (\Sexpr{round(sd(ans.summary$fig_class), 1)})\footnotemark[1]
             & \Sexpr{round(mean(ans.summary$vis_search), 1)} (\Sexpr{round(sd(ans.summary$vis_search), 1)})\\
Scaled Scores & 88.0 (\Sexpr{round(sqrt((24.6^2)*2), 1)})
              & 13.8 (4.5)
              & \Sexpr{round((120.0*294 + 114.9*323)/617/2, 1)} (\Sexpr{round(sqrt((30^2*294 + 27.8^2*323)/617)/2, 1)})\footnotemark[2]
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnotemark[3]
                & 13.8 (4.5)
                & \parbox[c]{.2\linewidth}{M:~120.0~(30.0)\\ F:~114.9~(27.8)}
                & --\\\hline
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.17\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.2\linewidth}{\footnotesize suburban 11th \& 12th \\ grade students\\(288-300 M, 317-329 F)}
              & \\\hline
\end{tabular}
\caption[Comparison of scores for cognitive tasks.]{Comparison of scores from Iowa State students and scores reported in~\protect\cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The visual search task~\protect\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment.
\label{tab:scorecomparison}}
\end{table*}

@@

Applying the grading protocol discussed in section~\ref{sec:scaling}, we see that the ranges of lineup and visuospatial test scores do not include zero; this indicates that we do not see random guessing from participants in any task. Figure~\ref{fig:Scores} shows the range of possible scores and the observed score distribution. 
Participants' scores on the VST indicate score compression; that is, both participants with medium and high visual search abilities scored at the extremely high end of the spectrum. 
In future experiments, participants should be given less time (or more questions) to better differentiate participants with medium and high-ability.
<<ResultsScaledScores, echo=F, include=F, fig.width=4.5, fig.height=4, warning=F, message=F>>=
loadfonts()
score.summary <- melt(ans.summary, measure.vars=21:25, value.name = "Score", variable.name = "Test")
score.summary$Test <- factor(score.summary$Test, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure Class.", "Paper Folding", "Visual Search"))
qrange$sqrt.k <- sqrt(qrange$k-1)
qrange$Test <- factor(qrange$testtype, levels=c("lineup", "card_rot", "fig_class", "folding", "vis_search"), labels=c("Lineups", "Card Rotation", "Figure Class.", "Paper Folding", "Visual Search"))

qr1 <- qrange
qr1$Type <- "Theoretical Range"
qr2 <- qrange
qr2$Type <- "Scores indicating guessing"
qr2$n <- with(qr2, -n/(k-1))

qrange <- rbind(qr1, qr2)
qrange$Type <- factor(qrange$Type)
qrange$Type <- factor(qrange$Type, levels=levels(qrange$Type)[2:1])

ggplot() + 
  geom_bar(data=qrange, aes(width=.9, x=Test, y=n, fill=Type), alpha=.35, stat="identity", position="identity", colour="red", inherit.aes=F) +
  scale_fill_manual("", values=c(NA, "red")) + 
  geom_boxplot(data=score.summary, aes(x=factor(Test), y=Score), inherit.aes=F, fill="grey") +
  theme_bw()  + 
  ylab("Test Scores") + 
  xlab("") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
  theme(legend.position=c(.68,.125),  legend.box.just = "left", 
        legend.key.height=unit(6, "mm")) 
  #coord_flip()
embedFonts("./Figure/fig-ResultsScaledScores-1.pdf")
@

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=9,fig.height=5.5, message=F, warning=F, fig.pos='ht'>>=
loadfonts()
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "STEM Major", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

tmp <- dlply(lineup.summary.categorical, .(variable), function(df){
    tmp <- anova(lm(lineup~factor(value, ordered=F), data=df))
    coefs <- as.data.frame(tmp)[1,]
    coefs$level <- gsub("factor(value, ordered = F)", unique(df$variable), rownames(coefs), fixed=T)
    rownames(coefs) <- NULL
    coefs <- coefs[,c(6, 1:5)]
    names(coefs) <- c("Variable", "DF", "Sum.of.Squares", "Mean.Squared", "F.value", "p.value")
    coefs
})

tmp <- rbind.fill(tmp)
tmp <- tmp[order(tmp$p.value, decreasing=F),]

lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, levels=unique(tmp$Variable))
lineup.summary.categorical$value <- factor(lineup.summary.categorical$value, levels=c("0","1", "2","3","4","5","[1, 2)", "[2, 5)", "5+", "18-20", "21+", "TRUE", "FALSE", "Female", "Male"))

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

@@

\subsection{Understanding Visual Abilities and Lineup Performance}
<<correlations, echo=F, comment="|", warning=F, message=F, results='asis', size="footnotesize", include=F>>=
print(xtable(cor(ans.summary[,c("lineup","card_rot", "fig_class", "folding", "vis_search")]), digits=c(1, rep(2, 5)), caption="Correlation matrix for the five tests.\\label{tab:corrmatrix5}"), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", scalebox=.9, table.placement="htb", caption.placement="top")
@

<<VisReasoningSPM,echo=FALSE, include=F, eval=T, fig.width=8,fig.height=8, message=F, warning=F>>=
data <- ans.summary[,c("lineup", "vis_search", "fig_class", "card_rot", "folding")]
grid <- expand.grid(x=1:ncol(data), y=1:ncol(data))
grid <- subset(grid, x != y & x<y)

all <- do.call("rbind", lapply(1:nrow(grid), function(i) {
  xcol <- grid[i, "x"]
  ycol <- grid[i, "y"]

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}
loadfonts()
require(grid)
ggplot()+
  geom_point(data=subset(all, points), aes(x=x, y=y)) + 
  geom_smooth(data=subset(all, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=subset(all, !points), aes(x=x, y=y, label=paste0("Correlation = \n", label))) + 
  geom_line(data=densities, aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank(),
        plot.margin=unit(c(0,0,0,0), unit="mm"))
rm(list=c("data", "all", "densities"))
embedFonts("./Figure/fig-VisReasoningSPM-1.pdf")
@
\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}

<<pca,echo=F, warning=F, message=F>>=
pca <- prcomp(ans.summary[,c("lineup", "card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
# screeplot(pca)
@

@@

A principal component analysis (PCA) of the four established visuo-spatial tests reveals that they all share a very strong first component, which explains about \Sexpr{round(summary(pca2)$importance[3,1],2)*100}\% of the total variability. Principal components (PC-) are ordered by importance (how much variability in the data they contain), and each principal component is uncorrelated with every other principal component.

PC1 is essentially an average across all tests representing a general ``visual intelligence" factor. The other principal components span another two dimensions, while the last dimension is weak (at \Sexpr{round(summary(pca2)$importance[2,4],2)*100}\%). 
PC2 differentiates the figure classification test from the visual searching test, whereas PC3 differentiates these two tests from the paper folding test. 
More detailed results from the 4-test analysis are provided in Appendix~\ref{app:pca}.

Incorporating the lineup task into the principal component analysis, we find the principal components to be fairly similar to the four-component analysis. 
Table~\ref{tab:PCAvariance5} shows the importance of each principal component. From the distribution of the variance components, we see that the lineup test  spans an additional dimension within the space of the four established tests. 
<<pcasummary5,echo=F, results='asis'>>=
print(xtable(summary(pca), caption="Importance of principal components, analyzing all five tests.\\label{tab:PCAvariance5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, size="footnotesize", table.placement="ht", caption.placement="top")
@

<<pcarotation5,echo=F, results='asis'>>=
print(xtable(pca$rotation, caption="PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \\label{tab:PCArotation5}",digits=2), sanitize.text.function=sanitize, sanitize.rownames.function=sanitize, sanitize.colnames.function=sanitize, table.placement="ht", caption.placement="top")
@
From the rotation matrix (see Table~\ref{tab:PCArotation5}) we see that  
the first principal component, PC1, is again essentially an average across all tests and accounts for \Sexpr{round(summary(pca)$importance[3,1]*100,1)}\% of the variance in the data. 
Biplots \cite{biplots:gabriel} of the remaining components are provided in Appendix~\ref{app:pca5}.

@@

% 
% The additional dimension, PC4 in the five-component PCA, separates lineups from the figure classification task and incorporates the relatively weak correlation between performance on the card rotation task and lineup performance; this dimension accounts for \Sexpr{round(summary(pca)$importance[2,4]*100,1)}\% of the variance. 
% PC5 is almost identical to PC4 in the four-component PCA; as before, it does not account for a significant portion of the variance in the data.
%

@@

A principal component of the five math/stats questions splits the variables into two main areas: the first principal component is an average of math skills, calculus 1 and STEM, while the second principal component is an average of having taken a statistics class and doing research. We therefore decided to use sums of these variables to come up with a separate math and a stats score. Note, that the correlation between the math and the stats score is almost zero. 
<<linearmodel, results='hide', echo=FALSE>>=
pca2 <- prcomp(ans.summary[,c("card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
ans.model <- data.frame(ans.summary, data.frame(pca2$x))
# exclude all variables that are not supposed to go into the linear model
# those are the unrotated tests; instead we have the PCs
excl <- c("id", "card_rot", "fig_class", "folding", "vis_search", "vidgame_hrs", "major1", "major2", "minor1", "minor2", "learning_disability", "colorblind", "epilepsy", "normal_vision")

ans.model$math_science_research <- ans.model$math_science_research=="y"
ans.model$stats_class <- ans.model$stats_class=="y"
ans.model$calc_1 <- ans.model$calc_1=="y"
pcmath <- prcomp(ans.model[,c("math_science_research", "stats_class", "calc_1", "math_skills", "stem")], scale=T)
pcmath$rotation
# first component of math PCA is average of math_skills, calc_1 and stem
# second component is stat_class and research

#biplot(pcmath)
#biplot(pcmath, 3:4)

xtabs(~stem+sex, data=ans.model)
# sad day!

xtabs(~stem+calc_1, data=ans.model)
# stem major is more associated with performance on lineups than calc 1...

xtabs(~stats_class+math_science_research, data=ans.model)
# only one kid is doing research without a stats class

# recode to sum of them
ans.model$STATS <- with(ans.model, stats_class + math_science_research)
ans.model$MATH <- with(ans.model, calc_1+math_skills/5+stem)

excl <- c(excl, c("math_science_research", "stats_class"))
excl <- c(excl, c("calc_1", "math_skills", "stem"))
incl <- setdiff(names(ans.model), excl)

m0 <- lm(lineup~., data=ans.model[, incl])

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
We fit a linear model of lineup scores in the thus modified demographic variables and the test scores from the visuo-spatial tests, selecting the best model using AIC and stepwise backwards selection. The result is shown in Table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as a measure of innate visual or intellectual ability, while the MATH score is a matter of both ability and training. The remaining principal components were not sufficiently associated with lineup score to be included in the model.

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous studies on different designs to investigate which plot type most effectively conveyed important characteristics of the underlying data set. 
Examples lineups for each experiment and more detailed explanations for each of these experiments are provided in Appendix~\ref{app:lineuptypes}.

<<LineupTypeSPM,echo=FALSE,include=F, eval=T, fig.width=4,fig.height=6.5, message=F, warning=F>>=
data <- lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "vis_search", "fig_class", "card_rot", "folding")]
grid <- expand.grid(x=1:ncol(data), y=1:ncol(data))
grid <- subset(grid, x != y & x<y)

all <- do.call("rbind", lapply(1:nrow(grid), function(i) {
  xcol <- grid[i, "x"]
  ycol <- grid[i, "y"]

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("Lineups_", "Lineup ") %>%
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Class.") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}

all2 <- subset(all, grepl("lineup", yvar) & points & !grepl("vis_search", xvar))
# names(all2)[1:4] <- c("yvar", "xvar", "y", "x")
all2mean <- rbind(
  ddply(subset(all2, !grepl("lineup", xvar)), .(xvar, yvar), summarize, label=paste0("rho == ", round(cor(x,y), 3)), x=min(x)-diff(range(x))/5, y=max(y)+diff(range(y))/5, hjust=0, vjust=1, size=3),
  ddply(subset(all2, grepl("lineup", xvar) & grepl("lineup", yvar)), .(xvar, yvar), function(df){
    with(df, data.frame(xvar=unique(yvar), yvar=unique(xvar), 
                        x=mean(range(y)), y=mean(range(x)), 
                        label=paste0("rho == ", round(cor(x,y),3)),
                        hjust=.5, vjust=.5, size=4))
    })
)
loadfonts()
ggplot()+
  geom_point(data=subset(all2, points), aes(x=x, y=y)) + 
  geom_smooth(data=subset(all2, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=all2mean, aes(x=x, y=y, label=label, hjust=hjust, vjust=vjust, size=size), parse=T) + 
  scale_size_identity() + 
  geom_line(data=subset(densities, grepl("lineup", xvar)), aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank())
embedFonts("./Figure/fig-LineupTypeSPM-1.pdf")
@

@@

<<removeAptitude,echo=F,fig.width=8, fig.height=4>>=
vistests <- scale(lineup.section.summary[,c("card_rot", "fig_class", "folding", "vis_search")])
lineups <- scale(lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3")])
vmain <- rowSums(vistests)/4
alltests <- data.frame(vmain, vistests-vmain, lineups)
pctests <- prcomp(alltests)
par(mfrow=c(1,2), mar=c(4,4.5,.1,.1), tcl=-0.01, cex.lab=.8,cex.axis=.7,mgp=c(2,.7,0),tcl=-.3)
#biplot(pctests)
#biplot(pctests, 3:4)
@

>>>>

\documentclass[journal]{vgtc}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\section{Results}\label{sec:results}
Results are based on an evaluation of 38 undergraduate students at Iowa State University. 61\% of the participants were in STEM fields, the others were distributed relatively evenly between agriculture, business, and the social sciences. Students were evenly distributed by gender, and were between 18 and 24 years of age with only one exception. This is reasonably representative\footnote{\url{http://www.registrar.iastate.edu/sites/default/files/uploads/stats/university/F14summary.pdf}} of the university as a whole; in the fall 2014 semester, 26\% of students were associated with the college of engineering, 24\% were associated with the college of liberal arts and sciences, 15\% were associated with the college of human sciences, 7\% with the college of design, 13\% with the business school, and 15\% with the school of agriculture.

@@

\begin{table*}[htb]\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure Classification & Visual Search  \\\hline
ISU Students & 83.4 (24.1) 
             & 12.4 (3.7)
             & 57.0 (23.8)\footnotemark[1]
             & 21.9 (2.3)\\
Scaled Scores & 88.0 (34.8)
              & 13.8 (4.5)
              & 58.7 (14.4)\footnotemark[2]
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnotemark[3]
                & 13.8 (4.5)
                & \parbox[c]{.2\linewidth}{M:~120.0~(30.0)\\ F:~114.9~(27.8)}
                & --\\\hline
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.17\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.2\linewidth}{\footnotesize suburban 11th \& 12th \\ grade students\\(288-300 M, 317-329 F)}
              & \\\hline
\end{tabular}
\caption[Comparison of scores for cognitive tasks.]{Comparison of scores from Iowa State students and scores reported in~\protect\cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The visual search task~\protect\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment.
\label{tab:scorecomparison}}
\end{table*}

@@

Applying the grading protocol discussed in section~\ref{sec:scaling}, we see that the ranges of lineup and visuospatial test scores do not include zero; this indicates that we do not see random guessing from participants in any task. Figure~\ref{fig:Scores} shows the range of possible scores and the observed score distribution. 
Participants' scores on the VST indicate score compression; that is, both participants with medium and high visual search abilities scored at the extremely high end of the spectrum. 
In future experiments, participants should be given less time (or more questions) to better differentiate participants with medium and high-ability.

@@

\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

@@

\subsection{Understanding Visual Abilities and Lineup Performance}

\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}

@@

A principal component analysis (PCA) of the four established visuo-spatial tests reveals that they all share a very strong first component, which explains about 64\% of the total variability. Principal components (PC-) are ordered by importance (how much variability in the data they contain), and each principal component is uncorrelated with every other principal component.

PC1 is essentially an average across all tests representing a general ``visual intelligence" factor. The other principal components span another two dimensions, while the last dimension is weak (at 6\%). 
PC2 differentiates the figure classification test from the visual searching test, whereas PC3 differentiates these two tests from the paper folding test. 
More detailed results from the 4-test analysis are provided in Appendix~\ref{app:pca}.

Incorporating the lineup task into the principal component analysis, we find the principal components to be fairly similar to the four-component analysis. 
Table~\ref{tab:PCAvariance5} shows the importance of each principal component. From the distribution of the variance components, we see that the lineup test  spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.1.2 by xtable 1.7-1 package
% Mon Mar 16 17:04:25 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.1.2 by xtable 1.7-1 package
% Mon Mar 16 17:04:25 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

From the rotation matrix (see Table~\ref{tab:PCArotation5}) we see that  
the first principal component, PC1, is again essentially an average across all tests and accounts for 60.1\% of the variance in the data. 
Biplots \cite{biplots:gabriel} of the remaining components are provided in Appendix~\ref{app:pca5}.

@@

% 
% The additional dimension, PC4 in the five-component PCA, separates lineups from the figure classification task and incorporates the relatively weak correlation between performance on the card rotation task and lineup performance; this dimension accounts for round(summary(pca)$importance[2,4]*100,1)\% of the variance. 
% PC5 is almost identical to PC4 in the four-component PCA; as before, it does not account for a significant portion of the variance in the data.
%

@@

A principal component of the five math/stats questions splits the variables into two main areas: the first principal component is an average of math skills, calculus 1 and STEM, while the second principal component is an average of having taken a statistics class and doing research. We therefore decided to use sums of these variables to come up with a separate math and a stats score. Note, that the correlation between the math and the stats score is almost zero.

We fit a linear model of lineup scores in the thus modified demographic variables and the test scores from the visuo-spatial tests, selecting the best model using AIC and stepwise backwards selection. The result is shown in Table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as a measure of innate visual or intellectual ability, while the MATH score is a matter of both ability and training. The remaining principal components were not sufficiently associated with lineup score to be included in the model.

% latex table generated in R 3.1.2 by xtable 1.7-1 package
% Mon Mar 16 17:04:25 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous studies on different designs to investigate which plot type most effectively conveyed important characteristics of the underlying data set. 
Examples lineups for each experiment and more detailed explanations for each of these experiments are provided in Appendix~\ref{app:lineuptypes}.

@@
